---
title: "NYPD Project"
author: "K.Crone"
date: "2023-09-24"
output: github_document

---

```{r, include=FALSE}
library(tidyverse)
library(dplyr)
library(ggplot2)
library(tinytex)
library(readr)
library(formatR)
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## NYPD Shooting Incidents Dataset

The dataset for this project comes from <https://catalog.data.gov/dataset/nypd-shooting-incident-data-historic>

The goal of this markdown project is to 1) Read in the dataset 2) Clean the data 3) Transform the data 4) Create Visualizations and Models to communicate the data

### Read in the data from the csv file

```{r reading file, tidy=TRUE}

nypd_data <- read.csv("NYPD_Shooting_Incident_Data__Historic_.csv",stringsAsFactors = FALSE)

summary(nypd_data) 
#view summary of the dataset as is before any cleaning

colnames(nypd_data) 
#view column names to make it easier to see what is needed and what is not needed for this dataset
```

### Cleaning & Transforming the Data

Several of the columns are not needed, so we will turn `nyp_data` into a dataframe, filter out the unwanted columns, clean up any missing/null/NA values, and change data types where needed

```{r clean_transform i, tidy=TRUE}
nypd_df <- data.frame(nypd_data)

str(nypd_df) 
# will show the data types of each column, 
#and the next variable will house the columns we want to keep

keep_cols <- c("OCCUR_DATE", "OCCUR_TIME", "BORO", "LOCATION_DESC","STATISTICAL_MURDER_FLAG","PERP_AGE_GROUP","PERP_SEX","PERP_RACE","VIC_AGE_GROUP", "VIC_SEX", "VIC_RACE")
new_nypd_df <- nypd_df[keep_cols]
```

```{r clean_transform ii, tidy=TRUE}
head(new_nypd_df)
tail(new_nypd_df)
colSums(is.na(new_nypd_df))
# check count of na values for each column
# LOCATION_DESC has by far the most na, and since that column is probably the least # valuable, we will remove it from out dataframe before cleaning out the rest of # the na's.
new_nypd_df = subset(new_nypd_df, select = -c(LOCATION_DESC) ) # remove column

new_nypd_df <- na.omit(new_nypd_df) # remove na values

summary(new_nypd_df) # we still retain 17,968 rows which seems like enough to use in the analysis

```

We have the columns we want, the na values are removed but we still need to check the data for any wrong types or weird entries/errors within each column

```{r clean_transform iii, tidy=TRUE}
unique(new_nypd_df['PERP_SEX']) 
# checking the PERP_SEX column reveals some null/unknown values still in the dataframe
new_nypd_df[new_nypd_df == "(null)"] = NA
new_nypd_df[new_nypd_df == "UNKNOWN"] = NA

ages <- c('<18', '18-24', '25-44', '45-64', '65+')
new_nypd_df <- new_nypd_df[new_nypd_df$PERP_AGE_GROUP %in% ages, ]
new_nypd_df <- new_nypd_df[new_nypd_df$VIC_AGE_GROUP %in% ages, ]


new_nypd_df <- na.omit(new_nypd_df)

unique(new_nypd_df['PERP_AGE_GROUP'])
unique(new_nypd_df$VIC_AGE_GROUP) 
# i am used to python , but it seems in R using the $ symbol for column names is the best way to call by column name...
```

The OCCUR_TIME might serve us better as a day/night flag, for our purposes, 6:00-18:00 will be daytime, and the rest is nighttime

```{r day_night}
# i spent a long time and tried to figure out the best way to do this but could not get it to work...

```

### Visualization & Analysis

```{r echo=FALSE, tidy=TRUE}
ggplot(new_nypd_df, aes(x = PERP_RACE, fill = PERP_AGE_GROUP)) +
  geom_bar(position = "stack") +
  labs(title = "Age Group Distribution of Perpetrators by Race", x = "Race of Perpetrator", y = "Count") +
  theme(
    text = element_text(size=10),
    plot.title = element_text(size=12, face='bold'),
    axis.title = element_text(size=10, face='bold'),
    axis.text = element_text(size = 9, angle = 45, hjust = 1),
    legend.title = element_text(face='bold')
    )


```

```{r echo=FALSE, tidy=TRUE}
boro_summary <- new_nypd_df %>%
  group_by(BORO) %>%
  summarize(Count = n())
ggplot(boro_summary, aes(x = "", y = Count, fill = BORO)) +
  geom_bar(width = 1, stat = "identity") +
  coord_polar(theta = "y") +  # This converts the bar chart to a pie chart
  theme_void()
```

```{r, tidy=TRUE}
# Let's format the date column so we can then get a visualization by year

new_nypd_df$OCCUR_DATE <- as.Date(new_nypd_df$OCCUR_DATE, format = "%m/%d/%Y")
new_nypd_df$Year <- year(new_nypd_df$OCCUR_DATE)

```

```{r echo=FALSE, tidy=TRUE}
# plotting the by-year graph

yearly_counts <- new_nypd_df %>%
  group_by(Year) %>%
  summarize(Count = n())

ggplot(yearly_counts, aes(x = Year, y = Count)) +
  geom_bar(stat = "identity") +
  labs(title = "Crimes per Year", x = "Year", y = "Number of Crimes") +
  theme_minimal()
```

### Model the Data

```{r}
# make the variables into factors 
new_nypd_df$STATISTICAL_MURDER_FLAG <- as.factor(new_nypd_df$STATISTICAL_MURDER_FLAG)
new_nypd_df$PERP_AGE_GROUP <- as.factor(new_nypd_df$PERP_AGE_GROUP)
new_nypd_df$PERP_RACE <- as.factor(new_nypd_df$PERP_RACE)
```

```{r}
#regression analysis
model <- glm(STATISTICAL_MURDER_FLAG ~ PERP_AGE_GROUP + PERP_RACE, 
             data = new_nypd_df, family = "binomial")

summary(model)
# the model seems to be indicating that race is not a significant factor, while age group is.

```
